---
title: "Pokemon and Their Strengths by Type, Stats, and Generation"
subtitle: "Proposal"
author: 
  - name: "Max Henderson"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Exploring whether a Pokemon's primary type predicts its attack and weight, and how legendary status and generation relate to total base stats."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
#| 
pokemon_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-01/pokemon_df.csv')

```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

Description: The R package provides English and Portuguese data, of 949 rows and 22 columns. Each row is one Pokemon speciesm identified by 'pokemon' (name) and 'species_id'. The columns include the physical measurements ('height' (m), 'weight' (kg)), the six core battle stats ('hp', 'attack', 'defense', 'special_attack', 'special_defense', 'speed'), 'base_experience', up to two type columns ('type_1', 'type_2'), color and egg-group metadata, image URLs, and 'generation_id'. I picked this dataset because I grew up with Pokemon shows, cards, and games, and ultimately would love to see if any of my own ideas about Pokemon trends are correct. I always thought 'fire' and 'fighting' types felt strong in design and stats.

```{r}
#| label: glimpse-data
glimpse(pokemon_df)
```

The dataset has no column that flags legendary or mythical Pokemon, even though that is a variable I need for Question 2. Both are addressed in the plan section.

## Questions

The two questions you want to answer.

1.  Is there a relationship between a Pokemon's primary type, its attack stat, and its weight? For example, do some types (like Fighting or Dragon) tend to be heavier and have higher attack?
2.  How do legendary status, generation, and total base stats (hp + attack + defense + special_attack + special_defense + speed) relate to each other? For example, do they stay consistent across generations or were they weaker or stronger for certain ID ranges? Is there a generational 'power creep'?

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

Question 1: As for wrangling, I will spot check for any issues as in zero or negatives, N/A's that may cause issues and address these. There shouldn't be more required cleaning than this.

For the two plots of Question 1, I intend to create a horizontal boxplot of 'attack' with 'type_1' reordered by median attack using 'fct_reorder()'. This will show the full spread with outliers that may drive a type's reputation, instead of just an average. For the second, I intend to create a scatterplot of 'weight' against 'attack'. This will include a facet_wrap(\~type_1) and a linear trend line, as it would be unfeasible to color all 18 types clearly.

Question 2: I will have to create a 'total_stats' variable which will be a mutate with the sum of the stats. I will also have to create a legendary/mythical flag, using the PokeAPI data. That dataset includes the boolean flag for is_legendary and is_mythical, so I can attach this to my data.

For the two plots of Question 2, I intend to create a boxplot of 'total_stats' by 'generation_id' with fill or facets split by 'legendary' to see whether legendaries separate from the pack within every generation or only some. For the second plot I intend to group by 'generation_id' and 'legendary' then summarize with a mean of the 'total_stats', to then plot a line chart with one line for legendary and one for standard Pokemon.

Task Name Status Due Priority Summary Data Wrangling In Prog. Week 1(this week) High Create variables etc. Connect PokeAPI To Do Week 1(this week) Moderate Find legendary flag Create Plots 1-4 To Do Week 2(next week) Moderate Work in RStudio\
Write up To Do Week 3 Low Complete writeup Presentation To Do Week 3 Low Create presentation
